---
title: "Market Segmentation for Airlines"
author: "David Fong"
date: "29th March 2019"
output: html_document
---

*Homework assignment for MITx "The Analytics Edge"*

Market segmentation is a strategy that divides a broad target market of customers into smaller, more similar groups, and then designs a marketing strategy specifically for each group. Clustering is a common technique for market segmentation since it automatically finds similar groups given a data set. 

In this problem, we'll see how clustering can be used to find similar groups of customers who belong to an airline's frequent flyer program. The airline is trying to learn more about its customers so that it can target different customer segments with different types of mileage offers. 

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(caret) # has preProcess to help normalize data
library(tidyverse)
library(ztable)
options(ztable.type = "html") # needed if outputting to HTML (knit)
```

## R Markdown

The file [AirlinesCluster.csv](https://prod-edxapp.edx-cdn.org/assets/courseware/v1/daa0a5e22087fe7ec788f56645d0302c/asset-v1:MITx+15.071x+1T2019+type@asset+block/AirlinesCluster.csv) contains information on 3,999 members of the frequent flyer program. This data comes from the textbook "Data Mining for Business Intelligence," by Galit Shmueli, Nitin R. Patel, and Peter C. Bruce. For more information, see the [website for the book](https://www.safaribooksonline.com/library/view/data-mining-for/9780470526828/).

```{r}
airlines <- read.csv("AirlinesCluster.csv")
str(airlines)
```

There are seven different variables in the dataset, described below:

    Balance = number of miles eligible for award travel
    QualMiles = number of miles qualifying for TopFlight status
    BonusMiles = number of miles earned from non-flight bonus transactions in the past 12 months
    BonusTrans = number of non-flight bonus transactions in the past 12 months
    FlightMiles = number of flight miles in the past 12 months
    FlightTrans = number of flight transactions in the past 12 months
    DaysSinceEnroll = number of days since enrolled in the frequent flyer program

```{r}
summary(airlines)
```


## Normalizing Data

If we don't normalize the data, the variables that are on a larger scale will contribute much more to the distance calculation, and thus will dominate the clustering.



```{r}
preproc <-preProcess(airlines) # pre-processes data
airlinesNorm <- predict(preproc, airlines) # performs normalization

summary(airlinesNorm)
```

All of the variables now have mean zero. Each of the variables also has standard deviation 1.

## Heirarchical clustering

```{r}
distance <- dist(airlinesNorm, method = "euclidean")
clusterairlines <- hclust(distance, method = "ward.D")
plot(clusterairlines)
```

If you run a horizontal line down the dendrogram, you can see that there is a long time that the line crosses 2 clusters, 3 clusters, or 7 clusters. However, it it hard to see the horizontal line cross 6 clusters. This means that 6 clusters is probably not a good choice.

## Five-cluster model

```{r}
clusterGroups <- cutree(clusterairlines, k=5) # five clusters

table(clusterGroups) # show the number of observations in each cluster

```

## Average values in each of the variables for the 5 clusters (the centroids of the clusters)

```{r results="asis"}
# results="asis" required to see HTML in 'knitted' document

spl <- split(airlines, clusterGroups) # splits data into subsets based on clusters (ClusterGroups)
# spl[[1]] is the first cluster group (same as "subset(movies[2:20], clusterGroups == 1)")

do.call(rbind, lapply(spl, colMeans)) %>% 
  ztable(caption = "Cluster centroids") %>%
  makeHeatmap(margin = 2)
# outputs the cluster centroid for all clusters
# 'lapply' runs the second argument (colMeans) on each element of the first argument (spl)
# do.call() applies 'rbind' to each of the lapply results
# ztable() creates the ztable.
# makeHeatmap() creates the heatmap. margin=2 creates a column-wise heatmap.
```

Cluster 1 mostly contains customers with few miles, but who have been with the airline the longest.

Cluster 2 contains customers with a large amount of miles, mostly accumulated through flight transactions.

Cluster 3 mostly contains customers with a lot of miles, and who have earned the miles mostly through bonus transactions.

Cluster 4 customers have the smallest value in DaysSinceEnroll, but they are already accumulating a reasonable number of miles. Mostly through non-flight transactions.

Cluster 5 customers have lower than average values in all variables. 

## k-means clustering

Creating five (5) clusters.

```{r}
set.seed(88)

k <- 5 # seven clusters

KMC <- kmeans(airlines, centers=k, iter.max = 1000)
str(KMC)
```

Cluster sizes

```{r}
KMC$size
```
## Average values in each of the variables for the 5 clusters (the centroids of the k-mean clusters)

```{r results="asis"}
# results="asis" required to see HTML in 'knitted' document

spl <- split(airlines, KMC$cluster) # splits data into subsets based on clusters (ClusterGroups)
# spl[[1]] is the first cluster group (same as "subset(movies[2:20], clusterGroups == 1)")

do.call(rbind, lapply(spl, colMeans)) %>% 
  ztable(caption = "Cluster centroids") %>%
  makeHeatmap(margin = 2)
# outputs the cluster centroid for all clusters
# 'lapply' runs the second argument (colMeans) on each element of the first argument (spl)
# do.call() applies 'rbind' to each of the lapply results
# ztable() creates the ztable.
# makeHeatmap() creates the heatmap. margin=2 creates a column-wise heatmap.
```

The clusters are not displayed in a meaningful order, so while there may be a cluster produced by the k-means algorithm that is similar to Cluster 1 produced by the Hierarchical method, it will not necessarily be shown first.

