---
title: "Analytical Policeman"
author: "David Fong"
date: "4th April 2019"
output: html_document
---

```{r setup, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(maps)
library(ggmap)
```

## data

We'll be using the motor vehicle theft dataset [mvt.csv](https://prod-edxapp.edx-cdn.org/assets/courseware/v1/52e7a3316f93ef4e5576fd6ff3ef0e98/asset-v1:MITx+15.071x+1T2019+type@asset+block/mvt.csv). Please download this dataset before starting this video. This data comes from the [Chicago Police Department](http://gis
.chicagopolice.org/). 

...and the [murders.csv](https://prod-edxapp.edx-cdn.org/assets/courseware/v1/8a471423a96f1f9f297b1d2545d077d3/asset-v1:MITx+15.071x+1T2019+type@asset+block/murders.csv) dataset from the U.S. Census Bureau and the FBI, as described [here](http://en.wikipedia.org/wiki/Gun_violence_in_the_United_States_by_state).

```{r}
mvt <- read.csv("mvt.csv", stringsAsFactors = FALSE)
str(mvt)
```

191641 observations of three variables.

Convert date variable to R format.

```{r}
mvt$Date <- strptime(mvt$Date, format="%m/%d/%y %H:%M")
mvt$Weekday <- weekdays(mvt$Date)
mvt$Hour <- mvt$Date$hour
str(mvt)
```

## Motor vehicle thefts by day of week

```{r}
WeekdayCounts <- as.data.frame(table(mvt$Weekday))
WeekdayCounts$Var1 <- factor(WeekdayCounts$Var1, ordered=TRUE,
                             levels = c("Monday", "Tuesday", "Wednesday",
                                        "Thursday", "Friday", "Saturday", "Sunday"))
# orders Var1 levels

str(WeekdayCounts)
```

```{r}
ggplot(WeekdayCounts, aes(x = Var1, y = Freq)) + 
  geom_line(aes(group = 1), linetype = 2, alpha = 0.3) +
  xlab("Day of the week") + ylab("Total motor vehicle thefts")
```

## Motor vehicle thefts. Day of week and hour

```{r}
DayHourCounts <- as.data.frame(table(mvt$Weekday, mvt$Hour))
DayHourCounts$Var1 <- factor(DayHourCounts$Var1, ordered=TRUE,
                             levels = c("Monday", "Tuesday", "Wednesday",
                                        "Thursday", "Friday", "Saturday", "Sunday"))
DayHourCounts$Hour <- as.numeric(as.character(DayHourCounts$Var2))
 # converts factor variable to a numeric variable
DayHourCounts
```

```{r}
ggplot(DayHourCounts, aes(x = Hour, y = Freq)) + geom_line(aes(group = Var1, color = Var1))
```

Heatmap

```{r}
ggplot(DayHourCounts, aes(x = Hour, y = Var1)) + geom_tile(aes(fill = Freq)) +
  scale_fill_gradient(name = "Total MV thefts", low = "white", high = "red") +
  theme(axis.title.y = element_blank()) # gets rid of y axis label
 # geom_tiles used for heatmap
```

## geographical heat map

Chicago map

```{r}
chicago <- ggmap::get_map(location = "chicago", zoom = 11)
ggmap(chicago)
```

The first one hundred points in the theft data.

```{r}
ggmap(chicago) + geom_point(data = mvt[1:100,], aes(x=Longitude, y = Latitude))
```

Crime counts data frame for each area

```{r}
LatLonCounts <- as.data.frame(table(round(mvt$Longitude, 2), round(mvt$Latitude, 2)))
 # rounded to two decimal places
str(LatLonCounts)
```

note that *Var1* and *Var2* are factors.

```{r}
LatLonCounts$Long <- as.numeric(as.character(LatLonCounts$Var1))
LatLonCounts$Lat <- as.numeric(as.character(LatLonCounts$Var2))
ggmap(chicago) + geom_point(data = LatLonCounts,
                            aes(x = Long, y = Lat,
                                color = Freq, size = Freq)) +
  scale_color_gradient(low = "yellow", high = "red")
```

```{r}

LatLonCounts2 <- subset(LatLonCounts, LatLonCounts$Freq>0)
 # removes observations with zero frequency theft, many of them in the sea!

ggmap(chicago) + geom_tile(data = LatLonCounts2,
                            aes(x = Long, y = Lat,
                                alpha = Freq), fill = "red")

```

## Murders and gun ownership in the United States

```{r}
murders <- read.csv("murders.csv")
str(murders)
```


51 observatioins for the fifty states (plus Washington DC). Six different variables : the name of the state, the population, the population density, the number of murders, the number of murders that used guns, and the rate of gun ownership.

```{r}
statesMap <- map_data("state")
ggplot(statesMap, aes (x = long, y = lat, group = group)) +
  geom_polygon(fill = "white", color = "black")
```

```{r}
str(murders$State)
```
```{r}
str(statesMap$region)
```

Note differences in the way that the two dataframes encode states (e.g. use of upper-case).

```{r}
murders$region <- tolower(murders$State)
murderMap <- merge(statesMap, murders, by = "region")
str(murderMap)
```

```{r}
ggplot(murderMap, aes(x = long, y = lat, group = group, fill = Murders)) +
  geom_polygon( color = "black") + 
  scale_fill_gradient(low = "black", high = "red", guide = "legend")
```

Most murders in California and Texas, but they are also the most populous states.

```{r}
ggplot(murderMap, aes(x = long, y = lat, group = group, fill = Population)) +
  geom_polygon( color = "black") + 
  scale_fill_gradient(low = "black", high = "red", guide = "legend")

```

```{r}
murderMap$MurderRate <- murderMap$Murders/murderMap$Population*100000

ggplot(murderMap, aes(x = long, y = lat, group = group, fill = MurderRate)) +
  geom_polygon( color = "black") + 
  scale_fill_gradient(low = "black", high = "red", guide = "legend")
```

The region with the highest murder rate is an outlier, Washington D.C., and it is also a very small region on the map that cannot be seen!

Create a new map where the limit is maximum 10.

```{r}
ggplot(murderMap, aes(x = long, y = lat, group = group, fill = MurderRate)) +
  geom_polygon( color = "black") + 
  scale_fill_gradient(low = "black", high = "red", guide = "legend",
                      limits = c(0, 10))
```

Gun ownereship

```{r}
ggplot(murderMap, aes(x = long, y = lat, group = group, fill = GunOwnership)) +
  geom_polygon( color = "black") + 
  scale_fill_gradient(low = "black", high = "red", guide = "legend")
```