---
title: "World Population 7,777,777,777"
subtitle: <h3>for ENVR-101 Introduction to Sustainability and Environmental Management</h3>
author: "David Fong"
date: "23rd March 2020"
abstract:
  When, according to currently available population data from www.worldometers.info, will the world population reach 7,777,777,777?
output: html_document
bibliography: "references.bib"
csl: apa.csl
---

<br />
<br />

## World population data

Population data derived from the World Population Clock website [@2020].

Dates and times are in Melbourne time, which up until 4th April 2020, is Australian Eastern Daylight Savings Time (AEDT).

```{r setup, message = FALSE, warnings = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lubridate) # for time functions
library(tidyverse)
library(chemCal) # for 'inverse.predict' function
library(formattable) # provide 'comma' function
```

```{r data}
d <- data.frame(Time = ymd_hms("2020-03-05 04:04:00", tz = "Australia/Melbourne"), Population = 7768625956)
d <- d %>% add_row(Time = ymd_hms("2020-03-06 21:40:00", tz = "Australia/Melbourne"), Population = 7769011917)
d <- d %>% add_row(Time = ymd_hms("2020-03-10 02:21:00", tz = "Australia/Melbourne"), Population = 7769723385)
d <- d %>% add_row(Time = ymd_hms("2020-03-11 04:23:00", tz = "Australia/Melbourne"), Population = 7769964915)
d <- d %>% add_row(Time = ymd_hms("2020-03-19 00:00:00", tz = "Australia/Melbourne"), Population = 7771705611)
d <- d %>% add_row(Time = ymd_hms("2020-03-20 06:51:00", tz = "Australia/Melbourne"), Population = 7771991830)
d <- d %>% add_row(Time = ymd_hms("2020-03-21 02:32:00", tz = "Australia/Melbourne"), Population = 7772174460)

d2 <- data.frame(Time = ymd_hms("2020-03-23 17:30:00", tz = "Australia/Melbourne"), Population = 7772758654)
d2 <- d2 %>% add_row(Time = ymd_hms("2020-03-25 07:52:00", tz = "Australia/Melbourne"), Population = 7773114617)
# observations not used in model creation, but will be used for model testing
```

## Population vs Time Plot

Population growth appears close to linear, as might be expected over a relatively short time period.

```{r plot}
ggplot(d, aes (x = Time, y = Population)) +
  geom_point() + # the data used for modeling
  geom_point(data = d2, aes (colour = "Test")) + labs(colour = "Dataset") 
```

## Normalizing and modifying the data

To help with model generation, a new variable **nTime** is calculated, where time '0' is defined as the time of the first observation (2020-03-05 04:04:00 AEDT). The unit of **nTime** is seconds.

Population growth is often modeled as exponential. **logPopulation** is defined as the natural logarithm of the population.

```{r}
d <- d %>%
  mutate(logPopulation = log(Population), # the log of the Population
         nTime = as.numeric(as.duration(Time - d$Time[[1]]))) 
# the first observation is time 'zero'

d2 <- d2 %>% # do the same for the test data
  mutate(logPopulation = log(Population), # the log of the Population
         nTime = as.numeric(as.duration(Time - d$Time[[1]]))) 
# the first observation is time 'zero'
```

## Models

### Model 1 - A simple linear model. 

Although population growth is often modeled as exponential, when using data from a relatively short time period, the growth might be approximated as linear.

$$
\begin{align}
Population &= \beta_0 + \beta_1 \times t\\
\end{align}
$$

where $t$ is the normalized time **nTime**.

```{r}
model1 <- lm(Population ~ nTime, data = d)
summary(model1)
```

### Model 2 - A simple exponential growth model.

$$
\begin{aligned}
Population & = \beta_0\times e^{\beta_1 \times t}\\
\implies log_e Population & = log_e \beta_0 + \beta_1 \times t
\end{aligned}
$$

```{r}
model2 <- lm(logPopulation ~ nTime, data = d)
summary(model2)
```

### Model 3 - A varying exponential growth model

Perhaps population growth is exponential, but the rate of exponential growth is changing with time e.g. declining due to demographic transition [@devries2013]. If, within the timeframe of the data, the rate of exponential growth change is locally linear, this can be modeled:

$$
\begin{aligned}
Population & = \beta_0\times e^{(\beta_1 + \beta_2 t ) t}\\
\implies log_e Population & = log_e \beta_0 + (\beta_1 + \beta_2 t) t\\
\implies log_e Population & = log_e \beta_0 + \beta_1t + \beta_2 t^2
\end{aligned}
$$


```{r}
model3 <- lm(logPopulation ~ nTime + I(nTime^2), data = d)
summary(model3)
```

## Model performance

```{r}
d2 # the test data
```

Are the predictions of the three models for (`r paste(d2$Time, collapse = " and ")`) close to the actual
observed population of (`r paste(comma(d2$Population, digits = 0), collapse = " and ")`)?

```{r}
# simple linear model
predict(model1, newdata = d2)
```
```{r}
# simple exponential growth model
exp(predict(model2, newdata = d2)) # need to raise e^logPopulation
```

```{r}
# varying exponential growth model
exp(predict(model3, newdata = d2)) # need to raise e^logPopulation
```

The predictions of the simple linear model (*model1*) and varying exponential growth model (*model3*) are both superior to the simple exponential growth model (*model3*), and almost identical.

But the simple linear model (*model1*) is the simpler model, requiring fewer variables, and so could be considered superior.


## Prediction for population 7,777,777,777

```{r}
target <- 7777777777
```


### Prediction plot

Population 'target' of `r comma(target, digits = 0)` indicated with the horizontal dashed line.

```{r warning=FALSE}

time_range <- ymd_hms("2020-03-05 04:04:00", tz = "Australia/Melbourne"):
  ymd_hms("2020-04-30 00:00:00", tz = "Australia/Melbourne")
time_range <- time_range[seq(1, length(time_range), 3600)] # only keep one 'x' value per hour
time_range <- as_datetime(time_range) # convert back to times

pred1 <- data.frame(Time = time_range,
                    Population = predict(model1,
                                         data.frame(nTime = as.numeric(time_range - time_range[1]))))
pred2 <- data.frame(Time = time_range,
                    Population = exp(predict(model2,
                                             data.frame(nTime = as.numeric(time_range - time_range[1])))))
pred3 <- data.frame(Time = time_range,
                    Population = exp(predict(model3,
                                             data.frame(nTime = as.numeric(time_range - time_range[1])))))


plotdata <- bind_rows(d %>% mutate(Observation = "Sample", Model = NA),
                      d2 %>% mutate(Observation = "Test", Model = NA),
                      pred1 %>% mutate(Model = "Model1", Observation = NA),
                      pred2 %>% mutate(Model = "Model2", Observation = NA),
                      pred3 %>% mutate(Model = "Model3", Observation = NA))

ggplot(plotdata, aes(x = Time, y = Population)) +
  geom_point(aes(shape = Observation)) + # the data used for modeling and testing
  scale_shape_manual(values = c(25,24), breaks = c("Sample", "Test")) +
  geom_line(aes(group = Model, color = Model)) +
  scale_color_manual(values = c("red", "green", "blue"),
                     breaks = c("Model1", "Model2", "Model3"),
                     guide = guide_legend(override.aes = list(shape = rep(NA, 3)))) +
  theme_light() +
  geom_hline(yintercept = target, color = "gold1", linetype = "dashed", size = 1) +
  expand_limits(x = ymd_hms("2020-04-30 00:00:00", tz = "Australia/Melbourne"))

```

All the model predictions for the time of world population achieving `r comma(target, digits = 0)` are in mid-April. In early April, Melbourne time changes from Daylight Savings (AEDT) to Australian Eastern Standard Time (AEST).

### Simple linear model '*model1*'

```{r}
x1 <- inverse.predict(model1, target)
prediction_time1 <- d$Time[1] + dseconds(x1$Prediction)
# dseconds 'duration' includes the change from Daylight Savings to Standard Time

x1
prediction_time1
with_tz(prediction_time1, tz = "GMT")
```

### Simple exponential growth model '*model2*'

```{r}
x2 <- inverse.predict(model2, log(target)) # need to 'log' the population
prediction_time2 <- d$Time[1] + dseconds(x2$Prediction)
# dseconds 'duration' includes the change from Daylight Savings to Standard Time

x2
prediction_time2
with_tz(prediction_time2, tz = "GMT")
```

### Varying exponential growth model '*model3*'


```{r}
# Can't use *inverse.predict*, because *inverse.predict* expects only one variable/predictor.

# Guess that the time, in seconds, is the same as the predicted by model1

x3 <- x1$Prediction

repeat {
  p <- round(exp(predict(model3, newdata = data.frame(nTime = x3))))
  if (p == target) {break} # reached the target!
  if (p < target) {x3 <- x3 + 1} # increment one second
  if (p > target) {x3 <- x3 - 1} # decrement one second
}

prediction_time3 <- d$Time[1] + dseconds(x3)
x3
prediction_time3
with_tz(prediction_time3, tz = "GMT")
```

## Results

The simple linear model prediction (*model1* : GMT `r with_tz(prediction_time1, tz = "GMT")`)  and varying exponential model prediction (*model3* : GMT `r with_tz(prediction_time3, tz = "GMT")`) are similar.

## Commentary

Although population growth is often modeled as exponential growth, the current human population in the world is very heterogeneous with regard to population growth. Some areas have a very youthful age structure, others are much older. Some areas have population growth which could be described as exponential growth, others have much more static, or even declining, population. If there are balancing influences on population growth, then population growth could be *logistic* [@devries2013], and currently in a linear growth phase.

### Model limitations

All three growth models base current growth on the entirety of the world population. In reality, only some age strata contribute to population growth. The size of these strata, and the reproductive rate of these strata, vary widely across the societies of the world.

Over the next few weeks www.worldometers.info might change their model to deal with unexpected influences on population, e.g. COVID-19 [@roser2020owid]!

# References

<div id = "refs"></div>