---
title: "Bikeshare"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(caTools)
library(ROCR)

library(rpart)
library(rpart.plot)
library(caret)
library(e1071)

```

## Load data


```{r}
bikes <- read.csv("bikes.csv")
summary(bikes)
```

In the dataset above, each observation represents one hour of the day (10886 hours). Here is a detailed description of the variables:

season: 1 = spring, 2 = summer, 3 = fall, 4 = winter 
holiday: whether the day is considered a holiday
workingday: whether the day is neither a weekend nor holiday
weather: 
       1: Clear, Few clouds, Partly cloudy, Partly cloudy

       2: Mist + Cloudy, Mist + Broken clouds, Mist + Few clouds, Mist 

       3: Light Snow, Light Rain + Thunderstorm + Scattered clouds, Light Rain + Scattered clouds 

       4: Heavy Rain + Ice Pallets + Thunderstorm + Mist, Snow + Fog 

temp: temperature in Celsius
atemp: "feels like" temperature in Celsius
humidity: relative humidity
windspeed: wind speed
count: number of total rentals
demand_level:1 if count is at least 250, 0 otherwise
hour: the hour of the day (0-23)

## Which season has the most rentals?


```{r}
bikes %>% group_by(season) %>% summarise(meancount = mean(count)) %>% arrange(desc(meancount))

bikes %>% group_by(season) %>% summarise(totalcount = sum(count)) %>% arrange(desc(totalcount))


```

## Creatining training set

```{r}
set.seed(100)

spl <- sample.split(bikes$demand_level, SplitRatio = 0.7) 
# 0.7 in training set, based on outcome variable $Reverse

train <-subset(bikes, spl == TRUE)
test <- subset(bikes, spl == FALSE)

count(train)

```

## Simple Logistic Regression

```{r}
simple_logit_model <- glm(demand_level ~ temp, data = train, family = "binomial")
summary(simple_logit_model)
```

Obtain predictions on the test set. Then, using a probability threshold of 0.5, create a confusion matrix for the test set.

```{r}

predicted <- predict(simple_logit_model, newdata = test, type = 'response')

accuracy.table <- table(test$demand_level, predicted >= 0.5)

accuracy.table

print(paste("Logit Model Accuracy : ", sum(diag(accuracy.table))/sum(accuracy.table)))
print(paste("Baseline accuracy :", 1-mean(test$demand_level)))


```
```{r}
print(paste("True positive rate : ", sum(test$demand_level & (predicted >= 0.5))/sum(test$demand_level)))
print(paste("False positive rate : ", sum(!(test$demand_level) & (predicted >= 0.5))/sum(!test$demand_level)))

```

### some correlations

```{r}
print(paste("Weather and temperature:", cor(train$weather, train$temp)))
print(paste("Weather and season:", cor(train$weather, train$season)))
print(paste("Working day and holiday:", cor(train$workingday, train$holiday)))
print(paste("Temp and atemp:", cor(train$temp, train$atemp)))
```

### another logit model

Train a logistic regression model now using all of the following variables in the training set:

season, holiday, workingday, weather, temp, humidity, windspeed, and hour

```{r}
logit_model <- glm(demand_level ~ season + holiday + workingday + weather + temp + humidity + windspeed + hour,
                   data = train, family = "binomial")
summary(logit_model)
```

Using new logistic regression model, obtain predictions on the test set. Then, using a probability threshold of 0.5, create a confusion matrix for the test set.

```{r}

predicted <- predict(logit_model, newdata = test, type = 'response')

accuracy.table <- table(test$demand_level, predicted >= 0.5)

accuracy.table

print(paste("Logit Model Accuracy : ", sum(diag(accuracy.table))/sum(accuracy.table)))

```

Plot the ROC curve for the logistic regression model.

```{r}
pred_ROCR <- prediction(predicted, test$demand_level)
auc_ROCR <- performance(pred_ROCR, measure = 'auc')
plot(performance(pred_ROCR, measure = 'tpr', x.measure = 'fpr'), colorize = TRUE,
     print.cutoffs.at = seq(0, 1, 0.1), text.adj = c(-0.2, 1.7))
```

```{r}
paste('Area under Curve :', signif(auc_ROCR@y.values[[1]]))
```

### Classification and Regression Tree

Use the caret package and the train function to perform 10-fold cross validation with the training data set to select the best cp value for a CART model that predicts the dependent variable demand_level using all of the possible independent variables except count which was used to define the dependent variable. Select the cp value from a grid consisting of the values 0.0001, 0.0002, 0.0003, ..., 0.02.

Remember to convert the demand_level column to a factor variable.

```{r}
set.seed(100)

numFolds <- trainControl(method = "cv", number = 10)
cpGrid <- expand.grid(.cp = seq(0.0001, 0.02, 0.0001))

train(as.factor(demand_level) ~ season+holiday+workingday+weather+temp+atemp+humidity+windspeed+hour,
      data = train, method = "rpart", trControl = numFolds, tuneGrid = cpGrid)

```

```{r}
treeCV <- rpart(as.factor(demand_level) ~ season+holiday+workingday+weather+temp+atemp+humidity+windspeed+hour,
                       data = train, method = 'class', cp = 0.001)

prp(treeCV)
```

```{r}
predictCV <- predict(treeCV, newdata = test, type = "class")

confusion_cv <- table(test$demand_level, predictCV)

confusion_cv
```


```{r}
print(paste("Accuracy:", sum(diag(confusion_cv))/sum(confusion_cv)))

```