---
title: "movie Lens"
author: "David Fong"
date: "3/28/2019"
output: html_document
---

Movie content clustering.

```{r setup, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ztable)
options(ztable.type = "html") # needed if outputting to HTML (knit)
```

## data

From [MovieLens website](http://files.grouplens.org/datasets/movielens/ml-100k/u.item).

Multiple movie genres. Each movie may have more than one genre.

```{r}
movies <- read.table("movieLens.txt", header = FALSE, sep = "|", quote="\"")
colnames(movies) = c("ID", "Title", "ReleaseDate", "VideoReleaseDate", "IMDB",
                    "Unknown", "Action", "Adventure", "Animation", "Childrens", "Comedy", "Crime",
                    "Documentary", "Drama", "Fantasy", "FilmnNoir", "Horror", "Music", "Mystery",
                    "Romance", "Scifi", "Thriller", "War", "Western")
str(movies)
```

Remove some variables which won't be used later.

```{r}
movies$ID <- NULL
movies$ReleaseDate <- NULL
movies$VideoReleaseDate <- NULL
movies$IMDB <- NULL
movies <- unique(movies)
str(movies)
```

How many movies are classified as comedies?

```{r}
table(movies$Comedy)
```

Westerns?
```{r}
table(movies$Western)
```
Romance AND drama?

```{r}
table(movies$Romance & movies$Drama)
```

## Create clusters

```{r}
distances <- dist(movies[2:20], method = "euclidean")
clusterMovies <- hclust(distances, method = "ward.D")
plot(clusterMovies)
```

Dendogram is difficult to read, because there are over 1,000 data ponts.

How many clusters should be picked? Application is to make movie recommendations
to users. Perhaps ten or so clusters. Hard to pick the right number of clusters
from the dendrogram.

## examine clusters

```{r}
clusterGroups <- cutree(clusterMovies, k=10) # ten clusters

cat("Average value of the actions variable for each of ten clusters:\n")
tapply(movies$Action, clusterGroups, mean) 
```

Lot of **action** movies in cluster 2, but none in cluster 4, 5, 7, 8, 9 or 10.

```{r}
cat("Average value of the actions variable for each of ten clusters:\n")
tapply(movies$Romance, clusterGroups, mean) 
```

All movies in cluster 6 and 7 are labelled **Romance**, whereas only 4% of movies cluster two are labeled as **romance** movies.

## cluster centroids

```{r results="asis"}
# results="asis" required to see HTML in 'knitted' document

spl <- split(movies[2:20], clusterGroups) # splits data into subsets based on clusters (ClusterGroups)
# spl[[1]] is the first cluster group (same as "subset(movies[2:20], clusterGroups == 1)")

do.call(rbind, lapply(spl, colMeans)) %>% 
  ztable(caption = "Cluster centroids") %>%
  makeHeatmap()
# outputs the cluster centroid for all clusters
# 'lapply' runs the second argument (colMeans) on each element of the first argument (spl)
# do.call() applies 'rbind' to each of the lapply results
# ztable() creates the ztable. makeHeatmap creates the heatmap
```

## cluster of specific movie "Men in Black"

```{r}
subset(movies, Title == "Men in Black (1997)")
cat("Men in Black belongs to cluster group : ", clusterGroups[257])
```

## some movies from cluster 2

```{r}
cluster2 <- subset(movies, clusterGroups == 2)
cluster2$Title[1:10]
```

## two cluster model

```{r results="asis"}
clusterGroups2 <- cutree(clusterMovies, k = 2)
spl <- split(movies[2:20], clusterGroups2) # splits data into subsets based on clusters (ClusterGroups)
# spl[[1]] is the first cluster group (same as "subset(movies[2:20], clusterGroups == 1)")

do.call(rbind, lapply(spl, colMeans)) %>% 
  ztable(caption = "Cluster centroids in 2-cluster model") %>% 
  makeHeatmap() 
# outputs the cluster centroid for all clusters
# 'lapply' runs the second argument (colMeans) on each element of the first argument (spl)
# do.call() applies 'rbind' to each of the lapply results
# ztable() creates the ztable. makeHeatmap creates the heatmap
```