---
title: "Predicting Stock Returns with Cluster-Then-Predict"
author: "David Fong"
date: "29th March 2019"
output: html_document
---

*Homework assignment for MITx "The Analytics Edge"*

Cluster-then-predict is a methodology in which you first cluster observations and then build cluster-specific prediction models. For example, this methodology can help improve the prediction of heart attack risk. We'll use cluster-then-predict to predict future stock prices using historical stock data.

When selecting which stocks to invest in, investors seek to obtain good future returns. We will first use clustering to identify clusters of stocks that have similar returns over time. Then, we'll use logistic regression to predict whether or not the stocks will have positive future returns.

```{r setup, message=FALSE, warn=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(knitr)
library(kableExtra)
library(formattable)
library(caTools) # training/test set creation tools
library(caret) # has preProcess command to normalize data
library(flexclust) # obtain training set and testing sut cluster assignments for observations
```

## data

For this problem, we'll use [StocksCluster.csv](https://prod-edxapp.edx-cdn.org/assets/courseware/v1/83c4f1cb4d1828257b6edf92c54d1f77/asset-v1:MITx+15.071x+1T2019+type@asset+block/StocksCluster.csv), which contains monthly stock returns from the NASDAQ stock exchange. The NASDAQ is the second-largest stock exchange in the world, and it lists many technology companies. The stock price data used in this problem was obtained from [infochimps](http://www.infochimps.com/datasets/nasdaq-exchange-daily-1970-2010-open-close-high-low-and-volume), a website providing access to many datasets.

Each observation in the dataset is the monthly returns of a particular company in a particular year. The years included are 2000-2009. The companies are limited to tickers that were listed on the exchange for the entire period 2000-2009, and whose stock price never fell below $1. So, for example, one observation is for Yahoo in 2000, and another observation is for Yahoo in 2001. Our goal will be to predict whether or not the stock return in December will be positive, using the stock returns for the first 11 months of the year.

```{r}
stocks <- read.csv("StocksCluster.csv")
str(stocks)
```

This dataset contains the following variables:

    ReturnJan = the return for the company's stock during January (in the year of the observation). 
    ReturnFeb = the return for the company's stock during February (in the year of the observation). 
    ReturnMar = the return for the company's stock during March (in the year of the observation). 
    ReturnApr = the return for the company's stock during April (in the year of the observation). 
    ReturnMay = the return for the company's stock during May (in the year of the observation). 
    ReturnJune = the return for the company's stock during June (in the year of the observation). 
    ReturnJuly = the return for the company's stock during July (in the year of the observation). 
    ReturnAug = the return for the company's stock during August (in the year of the observation). 
    ReturnSep = the return for the company's stock during September (in the year of the observation). 
    ReturnOct = the return for the company's stock during October (in the year of the observation). 
    ReturnNov = the return for the company's stock during November (in the year of the observation). 
    PositiveDec = whether or not the company's stock had a positive return in December (in the year of the observation). This variable takes value 1 if the return was positive, and value 0 if the return was not positive.

For the first 11 variables, the value stored is a proportional change in stock value during that month. For instance, a value of 0.05 means the stock increased in value 5% during the month, while a value of -0.02 means the stock decreased in value 2% during the month.

```{r}
summary(stocks)
```

54.61% of observations had positive returns in December.

April has the largest mean value (0.026308), and September has the smallest mean value (-0.014721).

## correlations

```{r}
data.frame(cor(stocks)) %>%
  formattable(list(area(col = 1:12, row = 1:12) ~ color_tile("#ffffff", "#d00404")))
```

The largest correlation coefficient between two return variables is 0.19167279, between ReturnOct and ReturnNov. 

## create training and testing sets

```{r}
set.seed(144)

spl <- sample.split(stocks$PositiveDec, SplitRatio = 0.7) # 70% in training set
stocksTrain <- subset(stocks, spl == TRUE)
stocksTest <- subset(stocks, spl == FALSE)
```

## logistic regression model

```{r}
StocksModel <- glm(PositiveDec ~ .,
                   data = stocksTrain, family = "binomial")

summary(StocksModel)
```

### logistic model prediction

Accuracy of the model on the *training* set.

```{r}
predictlog <- predict(StocksModel, newdata = stocksTrain, type = "response")

confusionmatrix <- table(predictlog > .5, stocksTrain$PositiveDec)
confusionmatrix

cat("\nAccuracy", sum(diag(confusionmatrix))/nrow(stocksTrain))
```

Accuracy of the model on the *test* set.

```{r}
predictlog <- predict(StocksModel, newdata = stocksTest, type = "response")

confusionmatrix <- table(predictlog > .5, stocksTest$PositiveDec)
confusionmatrix

cat("\nAccuracy", sum(diag(confusionmatrix))/nrow(stocksTest))
```

Accuracy on the *test* set of the *baseline model* that always predicts the most common outcome (PositiveDec = 1)

```{r}
mean(stocksTest$PositiveDec)
```

## clustering stocks

In cluster-then-predict, our final goal is to predict the dependent variable, which is unknown to us at the time of prediction. Therefore, if we need to know the outcome value to perform the clustering, the methodology is no longer useful for prediction of an unknown outcome value.

This is an important point that is sometimes mistakenly overlooked. If you use the outcome value to cluster, you might conclude your method strongly outperforms a non-clustering alternative. However, this is because it is using the outcome to determine the clusters, which is not valid. 

Remove the dependent variable before clustering.

```{r}
limitedTrain <- stocksTrain
limitedTrain$PositiveDec <- NULL
limitedTest <- stocksTest
limitedTest$PositiveDec <- NULL
```

Normalize data.

```{r}
preproc <- preProcess(limitedTrain)

normTrain <- predict(preproc, limitedTrain)
normTest <- predict(preproc, limitedTest)
```

```{r}
summary(normTrain)
```

```{r}
summary(normTest)
```

We see that the average (mean) return in January is slightly higher in the training set than in the testing set. Since normTest was constructed by subtracting by the mean ReturnJan value from the training set, this explains why the mean value of ReturnJan is slightly negative in normTest.

The distribution of the ReturnJan variable is different in the training and testing set.

### k-means clustering, three (3) clusters

```{r}
set.seed(144)

k <- 3
km <- kmeans(normTrain, centers = k)
str(km)
```

Cluster sizes

```{r}
table(km$cluster)
```

### Clustering stocks

flexclust package contains the object class KCCA, which stands for K-Centroids Cluster Analysis.

Need to convert the information from the clustering algorithm to an object of the class KCCA.
This conversion is needed before we can use the predict function on the test set tumorVector.

```{r}
km.kcca = as.kcca(km, normTrain)
clusterTrain = predict(km.kcca)
clusterTest = predict(km.kcca, newdata=normTest)
```

Test-set observations cluster assignment

```{r}
table(clusterTest)
```

Using the subset function, building data frames stocksTrain1, stocksTrain2, and stocksTrain3, containing the elements in the stocksTrain data frame assigned to clusters 1, 2, and 3, respectively 

Similarly build stocksTest1, stocksTest2, and stocksTest3 from the stocksTest data frame.

```{r}
stocksTrain1 <- subset(stocksTrain, clusterTrain == 1)
stocksTrain2 <- subset(stocksTrain, clusterTrain == 2)
stocksTrain3 <- subset(stocksTrain, clusterTrain == 3)

stocksTest1 <- subset(stocksTest, clusterTest == 1)
stocksTest2 <- subset(stocksTest, clusterTest == 2)
stocksTest3 <- subset(stocksTest, clusterTest == 3)
```

```{r}
stocksTrain %>%
  mutate(cluster = clusterTrain) %>%
  group_by(cluster) %>%
  summarise(mean(PositiveDec))
```

stocksTrain1 has the observations with the highest average value of the dependent variable. 

## cluster-specific logistic regression models

Build logistic regression models StocksModel1, StocksModel2, and StocksModel3, which predict PositiveDec using all the other variables as independent variables.

```{r}
StocksModel1 <- glm(PositiveDec ~ .,
                    data = stocksTrain1, family = "binomial")

summary(StocksModel1)
```

```{r}
StocksModel2 <- glm(PositiveDec ~ .,
                    data = stocksTrain2, family = "binomial")

summary(StocksModel2)
```

```{r}
StocksModel3 <- glm(PositiveDec ~ .,
                    data = stocksTrain3, family = "binomial")

summary(StocksModel3)
```

ReturnJan, ReturnFeb, ReturnMar, ReturnJune, ReturnAug, and ReturnOct differ in sign between the models. 

### cluster-specific predictions

```{r}
predictTest1 <- predict(StocksModel1, newdata = stocksTest1, type = "response") > .5

confusionmatrix1 <- table(predictTest1, stocksTest1$PositiveDec)
confusionmatrix1

cat("\nAccuracy", sum(diag(confusionmatrix1))/nrow(stocksTest1))
```

```{r}
predictTest2 <- predict(StocksModel2, newdata = stocksTest2, type = "response") > .5

confusionmatrix2 <- table(predictTest2, stocksTest2$PositiveDec)
confusionmatrix2

cat("\nAccuracy", sum(diag(confusionmatrix2))/nrow(stocksTest2))
```

```{r}
predictTest3 <- predict(StocksModel3, newdata = stocksTest3, type = "response") > .5

confusionmatrix3 <- table(predictTest3, stocksTest3$PositiveDec)
confusionmatrix3

cat("\nAccuracy", sum(diag(confusionmatrix3))/nrow(stocksTest3))
```

### overall cluster-specific prediction accuracy

```{r}
AllPredictions <- c(predictTest1, predictTest2, predictTest3)
AllOutcomes <- c(stocksTest1$PositiveDec, stocksTest2$PositiveDec, stocksTest3$PositiveDec)

confusionmatrix_all <- table(AllPredictions, AllOutcomes)
confusionmatrix_all

cat("\nAccuracy", sum(diag(confusionmatrix_all))/nrow(stocksTest))
```

We see a modest improvement over the original logistic regression model. Since predicting stock returns is a notoriously hard problem, this is a good increase in accuracy. By investing in stocks for which we are more confident that they will have positive returns (by selecting the ones with higher predicted probabilities), this cluster-then-predict model can give us an edge over the original logistic regression model.

