---
title: "Visualizing Network Data"
author: "David Fong"
date: "4th April 2019"
output: html_document
---

*assignment for "The Analytics Edge" (MITx)*

The cliche goes that the world is an increasingly interconnected place, and the connections between different entities are often best represented with a graph. Graphs are comprised of vertices (also often called "nodes") and edges connecting those nodes. 

```{r setup, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(igraph)
```

## data

We will visualize social networking data using anonymized data from Facebook; this data was originally curated in [a recent paper](http://i.stanford.edu/~julian/pdfs/nips2012.pdf) about computing social circles in social networks. In our visualizations, the vertices in our network will represent Facebook users and the edges will represent these users being Facebook friends with each other.

```{r}
edges <- read.csv("edges.csv")
users <- read.csv("users.csv")
```

The first file we will use, edges.csv, contains variables V1 and V2, which label the endpoints of edges in our network. Each row represents a pair of users in our graph who are Facebook friends. For a pair of friends A and B, edges.csv will only contain a single row -- the smaller identifier will be listed first in this row. From this row, we will know that A is friends with B and B is friends with A.

The second file, users.csv, contains information about the Facebook users, who are the vertices in our network. This file contains the following variables:

**id**: A unique identifier for this user; this is the value that appears in the rows of edges.csv

**gender**: An identifier for the gender of a user taking the values A and B. Because the data is anonymized, we don't know which value refers to males and which value refers to females.

**school**: An identifier for the school the user attended taking the values A and AB (users with AB attended school A as well as another school B). Because the data is anonymized, we don't know the schools represented by A and B.

**locale**: An identifier for the locale of the user taking the values A and B. Because the data is anonymized, we don't know which value refers to what locale.

```{r}
summary(users)

cat("\nNumber of users :", length(users$id))
```

```{r}
summary(edges)

cat("\nNumber of edges :", length(edges$V1))
```

We see that there are 146 pairs of users in our dataset who are Facebook friends. However, each pair (A, B) must be counted twice to count the total number of 'friendships', because B is a friend of A and A is a friend of B.

For our network, the average number of friends per user is 292/59=4.95.

Schools listed by students

```{r}
table(users$school)
```


Locale of students who list a school

```{r}
table(subset(users$locale, users$school != ""))
```

Gender vs school

```{r}
table(users$gender, users$school)
```

## visualizing the network

We will be using the igraph package to visualize networks.

A directed graph is one where the edges only go one way -- they point from one vertex to another. The other option is an undirected graph, which means that the relations between the vertices are symmetric. 
```{r}
g <- graph.data.frame(edges,FALSE, users)
 # edges is the data frame containing the symbolic edge list in the first two columns
 # FALSE - not a directed graph
 # users - vertex description

plot(g, vertex.size = 5, vertex.label = NA)
```

In this graph, there are a number of groups of nodes where all the nodes in each group are connected but the groups are disjoint from one another, forming "islands" in the graph. Such groups are called "connected components," or "components" for short. 

There are four components with at least two nodes in this graph.

There are seven users with no friends in the network.

In our graph, the "degree" of a node is its number of friends. We have already seen that some nodes in our graph have degree 0 (these are the nodes with no friends), while others have much higher degree. We can use degree(g) to compute the degree of all the nodes in our graph g.

```{r}
degree(g)
```

```{r}
cat("\nNumber of users with 10 or more Facebook users in this network :", sum(degree(g) >= 10))
```

In a network, it's often visually useful to draw attention to "important" nodes in the network. While this might mean different things in different contexts, in a social network we might consider a user with a large number of friends to be an important user. We know this is the same as saying that nodes with a high degree are important users.

To visually draw attention to these nodes, we will change the size of the vertices so the vertices with high degrees are larger. To do this, we will change the "size" attribute of the vertices of our graph to be an increasing function of their degrees:

```{r}
V(g)$size <- degree(g)/2 + 2

plot(g, vertex.label = NA)
 # we no longer user the vertex.size parameter when we plot the graph
```

## coloring the vertices

Thus far, we have changed the "size" attributes of our vertices. However, we can also change the colors of vertices to capture additional information about the Facebook users we are depicting.

When changing the size of nodes, we first obtained the vertices of our graph with V(g) and then accessed the the size attribute with *V(g)$size*. To change the color, we will update the attribute *V(g)$color*.

To color the vertices based on the gender of the user, we will need access to that variable. When we created our graph g, we provided it with the data frame users, which had variables gender, school, and locale. These are now stored as attributes *V(g)$gender*, *V(g)$school*, and *V(g)$locale*.

We can update the colors by setting the color to black for all vertices, than setting it to red for the vertices with gender A and setting it to gray for the vertices with gender B:

```{r}
V(g)$color <- "black"
V(g)$color[V(g)$gender == "A"] <- "red"
V(g)$color[V(g)$gender == "B"] <- "gray"

plot(g, vertex.label = NA)
```

## coloring vertices by school

```{r}
V(g)$color <- "black"
V(g)$color[V(g)$school == "A"] <- "red"
V(g)$color[V(g)$school == "AB"] <- "gray"

plot(g, vertex.label = NA)
```

## coloring vertices by locale

```{r}
V(g)$color <- "black"
V(g)$color[V(g)$locale == "A"] <- "red"
V(g)$color[V(g)$locale == "B"] <- "gray"

plot(g, vertex.label = NA)
```

```{r}
```
