---
title: "WHO plots"
author: "David Fong"
date: "4/3/2019"
output: html_document
---

*Exercise in the "The Analytics Edge" (MITx)*

```{r setup, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

## data

```{r}
WHO <- read.csv("WHO.csv")
str(WHO)
```

194 observations (countries) for 13 different variables

* **Country** name of country
* **Region** region
* **Population** in 1000s
* **Under15** percentage under 15
* **Over60** percentage over 60
* **FertilityRate** average number of children per woman
* **LifeExpectancy** in years
* **ChildMortality** number of children who die by age five per 1000
* **CellularSubscribers** Number of cellular ubscribers per 100
* **LiteracyRate** Literacry rate among adults older than 15 years
* **GNI** Gross national icome per capita
* **PrimarySchoolEnrollmentMale** percentage of male children enrolled in primary school
* **PrimarySchoolEnrollmentFemale** percentage of female children enrolled in primary school

## Plots

### Fertility Rate vs GNI 

Using basic plot package

```{r}
plot(WHO$GNI, WHO$FertilityRate)
```

Using ggplot2

```{r}
scatterplot <- ggplot(WHO, aes(x = GNI, y = FertilityRate))
fertilityGNIplot <- scatterplot + 
  geom_point(color = "blue", size = 3, shape = 17) + 
  ggtitle("Fertility Rate vs Gross National Income")

fertilityGNIplot
```

Save the fertility vs GNI plot into a PDF

```{r}
pdf("MyPlot.pdf")
print(fertilityGNIplot)
dev.off() # closes the pdf file
```

color points by region

```{r}
ggplot(WHO, aes(x=GNI, y=FertilityRate, color=Region)) + geom_point()
```

Points from regions are located in different areas of the plot.

Now colour according to life expectancy.

```{r}
ggplot(WHO, aes(x=GNI, y=FertilityRate, color=LifeExpectancy)) + geom_point()
```


```{r}
ggplot(WHO, aes(x=FertilityRate, y=Under15, color=Region)) + geom_point() + scale_color_brewer(palette="Dark2")
```

Appears to be a correlation, but as the fertility rate increases, the Under15 variable increases less.

Not a linear relatioinship, but perhaps a log transformation of FertilityRate will be better.

Graph Under 15 vs log(FertilityRate)

```{r}
ggplot(WHO, aes(x=log(FertilityRate), y=Under15)) + geom_point()
```

### Create a model

```{r}
model <- lm(Under15 ~ log(FertilityRate), data=WHO)
summary(model)
```

Highly significant, with $R^2$ 0.9387

compare this to a model which *doesn't* use the log transformation.

```{r}
model <- lm(Under15 ~ FertilityRate, data=WHO)
summary(model)

```

$R^2$ is less, at 0.8756

Add a regression line to the Under15 vs log(FertilityRate) plot

```{r}
ggplot(WHO, aes(x=log(FertilityRate), y=Under15)) + geom_point() + stat_smooth(method="lm")
```

Regression line with 95% confidence interval.

Can change to 99% confidence interval.

```{r}
ggplot(WHO, aes(x=log(FertilityRate), y=Under15)) + geom_point()+ 
  stat_smooth(method="lm", level=0.99)
```

or with no confidence interval at all

```{r}
ggplot(WHO, aes(x=log(FertilityRate), y=Under15)) + geom_point()+ 
  stat_smooth(method="lm", se=FALSE, color="orange")
```